---
title: "Prosessing"
format: html
---

### Importing Packages

```{r}
#| message: false
library(tidyverse)
library(here)
library(readr)
```

### Medicine Equality in Education

first we load the cleaned data and filter out all patients that haven't registered education.

```{r}
#| message: false
med_edu <- read_csv(here("data/med_data_tidy.csv")) |> filter(!is.na(edu))
med_edu 
```

Getting an overview of the amount of years of education shows that 8-15 years is the most common and will heavily scew the coming analysis.

```{r}
med_edu |> ggplot(mapping = aes(x = edu)) + 
  geom_bar() + 
  labs(title = "Education frequency",        
    center =TRUE) +
  xlab("Years of education") + ylab("People")

```

We do not have enough datapoints to have a meaningful death rate for all amounts of education, so we chose to split them up in intervals.

```{r}
med_edu <- med_edu |> 
  mutate(edu = case_when(0 <= edu & edu < 9 ~ "(0-9]",
                         9 <= edu & edu < 18 ~ "(9-18]",
                         18 <= edu & edu < 30 ~ "(18-30]"), 
         edu = factor(edu, levels = c("(0-9]","(9-18]", "(18-30]"))) |> 
  mutate("low" = case_when(edu == "(0-9]" ~ 1,
                           .default = 0)) |>
  mutate("medium" = case_when(edu == "(9-18]" ~ 1,
                           .default = 0)) |>
  mutate("high" = case_when(edu == "(18-30]" ~ 1,
                           .default = 0))

med_edu |> write_rds(here("data/df_educat.RDS"))

med_edu 
```

```{r}
med_edu |>
  filter(!is.na(edu)) |>
  ggplot(aes(x = edu, y = death)) +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun = mean, geom = "line", group = 1, color = "black") +
  geom_hline(yintercept = mean(med_edu$death, na.rm = TRUE), 
             linetype = "dotted", color = "red", linewidth = 1)

```

```{r}
edu_death_model <- lm(death ~ edu, data = med_edu)
edu_death_model |> summary()
```

```{r}

med_edu_low_model <- lm(death ~ low, data = med_edu)
med_edu_low_model |> summary()
```

```{r}


med_edu_medium_model <- lm(death ~ medium, data = med_edu)
med_edu_medium_model |> summary()
```

```{r}

med_edu_high_model <- lm(death ~ high, data = med_edu)
med_edu_high_model |> summary()
```

```{r}
df_med_edu_models <- tibble(group = c("edu_low", "edu_medium", "edu_high"),
                            model = list(med_edu_low_model, 
                                      med_edu_medium_model, 
                                      med_edu_high_model))

df_med_edu_models |> write_rds(here("data/models/edu.RDS"))
```
